Seniority and Hierarchy Configuration Interaction for Radicals and \\ Excited States

### 2023-10-25 Manuscript/Response_Letter/Response_Letter.tex

{Figures 3 and 4: The data is shown with respect to the number of determinants, which makes the
reader to evaluate the different methods efficiently. However, for readability, I think it will also be
helpful to provide n in eCIn, sCIn, and hCIn in these figures.
}

\alert{
We have experimented including the values of $n$ to the figures, as suggested by the reviewer.
However, we consider that they become somewhat cluttered and less readable, and therefore we prefer to leave them unchanged.
With the intention of helping the reader, we have supplemented the following sentence to the captions of Figures 3 (4):
``Each point denotes one CI model, according to the sequences: HF, hCI1, hCI1.5, hCI2, etc. (green); HF, CIS, CISD, etc. (red); and sCI0, sCI2, and sCI4 (sCI1, sCI3, sCI5) (blue).''
}

{In sections IV-C and IV-D: The EN2 corrections are not discussed at all in the excited states, making
it unnatural compared to the presentation of the radical systems. While it might be challenging to
perform the additional calculations for all the systems, I think it will be helpful for some sample
systems to investigate the impact of EN2 correction in hCI and DhCI calculations (maybe in the
same fashion presented in Figures 3 and 4).
}

\alert{
We agree with the reviewer's suggestion that including the EN2 correction for the excited state calculations is a natural and important addition to the manuscript.
We have performed additional calculations for the full set of states with the different CI models surveyed here (except for the more computationally demanding higher-order models).
These new results were included in Tables II, III, and in the Supporting Information, while a new subsection (IV.E), starting on page 13, is devoted to their discussion.
Additions to the Conclusion, on page 14, and to the abstract have also been made.
}

@@

{The authors present a generalization of the hierarchy configuration interaction (hCI) method to an arbitrary reference state, which allows them to extend their previously developed hCI model to open-shell systems and general excited-state electronic structures. They test the performance of their proposed extension against various test sets containing (small) molecules for which reliable reference data is available. Their numerical analysis and discussion are in-depth, while their methodology (the theory part) is comprehensive. Their numerical study allows them to obtain reliable energies and spectroscopic properties for their more elaborate hCI models. This manuscript is of potential interest to the readership of the Journal of Chemical Theory and Computation. However, a few remarks need to be addressed before publication.
}

\alert{
We thank the reviewer for the positive appreciation of our work.
}

{1) The sentence ``[...] whose number $\Ndet$ share the same scaling with the number of basis functions $N$ [...]'' is rather difficult to understand. The authors need to either rephrase or elaborate more.
}

\alert{
This sentence has been rephrased and now reads, on page 2,
``The number of determinants $\Ndet$ within a given class scales polynomially with the number of basis functions $N$, with the exponent depending on the specific class.
For instance, for the class of doubly-excited determinants with no unpaired electrons ($e=2$ and $s=0$), $\Ndet = \order*{N^2}$.
hCI was defined such that, at a given hierarchy $h$, all classes of determinants presenting a scaling of $\Ndet = \order*{N^{2h}}$ or less are accounted for.''
}

@@

{2) the authors need to include more information concerning the fitting procedure in the main manuscript for reproducibility reasons. In the SI, they state that they used a Morse function. Why did the authors use a Morse function and not, for instance, a generalized Morse function, etc... Furthermore, it is unclear how they fitted their vibrational frequencies.
}

\alert{
The reviewer's questions are addressed on page 3 of the revised Supporting Information, reading
``By fitting the computed potential energy curves with a Morse potential $V(r) = D [1 - e^{-a(r-r_e)}]^2 + C $,
we obtained equilibrium geometries $r_e$, force constants $2Da^2$, and harmonic vibrational frequencies $\left( Da^2/2\pi^2\mu \right)^{1/2}$,
where $\mu$ is the reduced mass.
Since the fitting is restricted to the equilibrium region, employing more elaborate functions would be inconsequential, such that the adopted Morse potential is well justified.
The fitting was performed with the Marquardt-Levenberg algorithm implemented in Gnuplot.''
%Since the fitting is restricted to the equilibrium region (the ranges are given in the Supporting Information),
%using a more elaborate functional form would only produce minimal changes to the obtained fitted values. A Morse potential is thus enough for our purposes.
%Moreover, any residual bias from the fitting function would be inconsequential to the goal of comparing to the (equally fitted) FCI values.
We prefer to keep these details in the Supporting Information rather than in the main text.
}

@@

{3) Some of the PES displayed in the SI are not smooth and rather kinky. How does the lack of smoothness affect the fitting procedure and, hence, the vibrational frequencies?
}

\alert{
\ce{CN} is the only system for which some of the potential energy curves are not smooth around the equilibrium geometry.
This is mentioned on page 6 of the revised manuscript, reading
``For the \ce{CN} radical, CIS, sCI1, and sCI3 produce crossings between ground and excited states around equilibrium, hence non-smooth adiabatic PECs
and somewhat less reliable equilibrium properties in these specific cases.''
}

@@

\clearpage

{4) The authors need to mention (or draw) the molecules of their excited state study. Not everyone might be familiar with the QUEST database. It would be insightful to mention the character of the excited states. I assume these are all singly excited states (no double excitation character?). Can the authors provide additional information regarding their nature (local excitation, charge transfer, Rydberg, etc.)? Specifically, what kind of excited states can be described well or less well with hCI methods?
}

\alert{
The set of systems we considered is now presented in Table II, whereas on page 5 we describe the character of the excited states, reading
``Our set includes mostly small systems, ranging from \ce{BH} to glyoxal, displaying a mix of valence and Rydberg singly-excited states,
and 2 doubly-excited states (glyoxal and nitroxyl). It does not include large molecules nor charge transfer states.''
The details on each specific excited state can be found in The Supporting Information.
The revised manuscript now mentions the performance of the hCI models for different types of excited states, on page 11:
``$\Delta$hCI2 is more accurate for singlets than for triplets (MAEs of \SI{0.16}{\eV} and \SI{0.25}{\eV}),
whereas Rydberg states are better described than valence states (MAEs of \SI{0.14}{\eV} and \SI{0.24}{\eV}).
The same trends are found for the low-order models ($\Delta$hCI1 and $\Delta$hCI1.5), and also for $\Delta$CISD, which slightly outperforms $\Delta$hCI2 for each type of excitation.
The specific MAEs can be found in the Supporting Information.''
This aspect is also discussed for the hCI+EN2 models in the new subsection (IV.E).
}

@@

{5) For reasons of transparency, the authors should define MSE, MAE, RMSE, and SDE using equations, even if it might be obvious.
}

\alert{
The statistical measures are now defined on page 2 of the Supporting Information, which are referred to on page 9 of the main text, reading
``For completeness, the definition of these statistical measures can be found in the Supporting Information.''
}

@@

{6) The authors keep mentioning ``reference theoretical values.'' Please mention what kind of reference values were used.
}

\alert{
This information is provided in the Supporting Information, which is stated on page 5 of the revised manuscript, reading
``For each excited state considered here, the reference value and the corresponding method to obtain it (of high-order coupled cluster or extrapolated FCI quality) can be found in the Supporting Information.''
}

@@

{7) The authors look at several small-sized molecules for which FCI can still be determined. There is nothing wrong with that. However, I am missing numerical studies on slightly larger systems, which are known to be challenging in computational chemistry. I recommend adding some ``larger'' molecules, which feature open- and closed-shell ground and excited states. Examples are cyclobutadiene (its automatization problem) and o-, m-, and p-benzynes. These systems are notoriously difficult to study and extensively discussed in the literature. Including just those four molecules would add more value to the paper and allow a direct comparison to other state-of-the-art computational methods.
}

\alert{
This is a pertinent remark.
From the examples suggested by the reviewer, we decided to investigate the automerization problem of cyclobutadiene.
Our results can be found in a new paragraph on page 9, starting as:
``We further explored hCI to describe the automerization barrier in the ground state of cyclobutadiene...''
On the other hand, the case of benzynes is left as a perspective for the future, otherwise, the manuscript would be excessively lengthy in our opinion.
Moreover, our code would need further optimizations in order to tackle efficiently these larger systems.
We notice that extensive CI+EN2 additional calculations have been carried out and a corresponding new subsection (IV.E) has been added to the manuscript, based on a suggestion from Reviewer \#1.
We further highlight the point raised by Reviewer \#2 in the Conclusion, reading
``hCI also produced reasonable energies for the automerization barrier of cyclobutadiene.
It remains to be seen how these models perform for other strongly correlated systems, larger molecules, and charge transfer excited states.''
}

### 2023-10-25 Manuscript/manuscript.tex
ready to submit

% Abstract
\begin{abstract}
Hierarchy configuration interaction (hCI) has been recently introduced as an alternative configuration interaction (CI) route combining excitation degree and seniority number, which showed to efficiently recover both dynamic and static correlations for closed-shell molecular systems [\href{https://doi.org/10.1021/acs.jpclett.2c00730}{\textit{J.~Phys.~Chem.~Lett.}~\textbf{2022}, \textit{13}, 4342}].
Here, we generalize hCI for an arbitrary reference determinant, allowing calculations for radicals and for excited states in a state-specific way.
We gauge this route against excitation-based CI (eCI) and seniority-based CI (sCI) by evaluating how different ground-state properties of radicals converge to the full CI limit.
We find that hCI outperforms or matches eCI, whereas sCI is far less accurate, in line with previous observations for closed-shell molecules.
Employing the second-order Epstein-Nesbet \fk{(EN2)} perturbation theory as a correction significantly accelerates the convergence of hCI and eCI.
We further explore various hCI and sCI models to calculate excitation energies of closed- and open-shell systems.
Our results underline that both the choice of the reference determinant and the set of orbitals drive the fine balance between correlation of ground and excited states.
State-specific hCI2 and higher order models perform similarly to their eCI counterparts, whereas lower orders of hCI deliver poor results,
\fk{unless supplemented by the EN2 correction, which substantially improves their accuracy.}
In turn, sCI1 produces decent excitation energies for radicals, encouraging the development of related seniority-based coupled-cluster methods.
\bigskip
\begin{center}
        \boxed{\includegraphics[keepaspectratio,width=3.25in]{TOC}}
\end{center}
\bigskip
\end{abstract}

@@

Here, our first goal is to generalize hCI for an arbitrary type of reference, thus extending its applicability from ground-state closed-shells \cite{Kossoski_2022}
to radicals and state-specific excited-state calculations.
This is done in Sec.~\ref{sec:hCI}.
The computational details about the implementation and the specific calculations performed here are presented in Sec.~\ref{sec:compdet}.
In Sec.~\ref{sec:res_A}, we assess the performance of hCI, eCI, and sCI models by calculating various properties of four ground-state radicals, which comprises our second goal.
Our third goal is to evaluate how perturbation theory, more precisely the second-order Epstein-Nesbet (EN2) perturbative correction (both standard and renormalized), \cite{Garniron_2019}
impacts various properties of ground-state closed-shell systems and radicals, for hCI, eCI, and sCI models.
This part is presented in Sec.~\ref{sec:res_B}.
Inspired by the results of hCI for ground-state closed-shell systems, \cite{Kossoski_2022}
and by the promising set of excitation-based $\Delta$CI, \cite{Kossoski_2023} here we explore hCI models for excited states, following both the ground-state-based and state-specific approaches.
In this sense, our fourth goal is to assess the accuracy of hCI models for excited states, which is the subject of Sec.~\ref{sec:res_C}.
Furthermore, to the best of our knowledge, sCI models have not yet been directly used to target excited states,
despite the growing number of methods that exploit the concept of seniority,
for both ground
\cite{Limacher_2013,Limacher_2014,Tecmer_2014,Boguslawski_2014a,Boguslawski_2015,Boguslawski_2014b,Boguslawski_2014c,Johnson_2020,Henderson_2014,Stein_2014,Henderson_2015,Chen_2015,Bytautas_2018,Marie_2021,Boguslawski_2021,Tecmer_2022,Mamache_2023,Fecteau_2022}
and excited states.
\cite{Boguslawski_2016b,Boguslawski_2016c,Boguslawski_2019,Nowak_2019,Kossoski_2021,Marie_2021,Tecmer_2022,Rishi_2023,Nowak_2023,Fecteau_2022}
Our fifth goal, detailed in Sec.~\ref{sec:res_D}, is therefore to define and gauge ground-state-based and state-specific sCI models for excited states.
\fk{Finally, in Sec.~\ref{sec:res_E}, we assess how the excitation energies are impacted by the EN2 perturbative correction to hCI and sCI models, our sixth and last goal.}
Section \ref{sec:conclusion} closes the present contribution with the main conclusions and perspectives.

@@

We introduced hCI \cite{Kossoski_2022} as a particular truncation of the Hilbert space,
viewed as a two-dimensional map of determinants built from their seniority $s$ and their excitation degree $e$ with respect to a reference determinant, as shown in Fig.~\ref{fig:hCI}.
By defining orbital subspaces according to the occupancy in the reference determinant (doubly-occupied, singly-occupied, or unoccupied),
we refer to the class of an excited determinant as the combination of the number of electrons and seniority in each orbital subspace.
%hCI was defined such that all classes of determinants whose number $\Ndet$ share the same scaling with the number of basis functions $N$ are included at the same order of truncation $h$,
\fk{The number of determinants $\Ndet$ within a given class scales polynomially with the number of basis functions $N$, with the exponent depending on the specific class.
For instance, for the class of doubly-excited determinants with no unpaired electrons ($e=2$ and $s=0$), $\Ndet = \order*{N^2}$.
hCI was defined such that, at a given hierarchy $h$, all classes of determinants presenting a scaling of $\Ndet = \order*{N^{2h}}$ or less are accounted for.}
This means moving diagonally in the seniority-excitation map (denoted by the color tones in Fig.~\ref{fig:hCI}).
In comparison, eCI spans the map horizontally (top to bottom), whereas sCI does it vertically (left to right).

@@

The various CI models introduced here were further assessed based on calculated vertical excitation energies for 69 electronic states,
from 17 closed-shell systems and 13 radicals \fk{(shown in Table~\ref{tab:systems})}, with geometries extracted from the QUEST database. \cite{Veril_2021}
\fk{Our set includes mostly small systems, ranging from \ce{BH} to glyoxal, displaying a mix of valence and Rydberg singly-excited states,
and 2 doubly-excited states (glyoxal and nitroxyl). It does not include large molecules nor charge transfer states.}
The full set of excited states and calculated excitation energies, for the various CI models, are provided in the {\SupInf}.

\begin{table}[ht!]
\caption{\fk{Systems Considered in the Excited State Calculations.}}
\label{tab:systems}
\begin{ruledtabular}
\begin{tabular}{llllll}
\mc{3}{c}{Closed-shells}           & \mc{3}{c}{Open-shells} \\
\cline{1-3} \cline{4-6}
glyoxal      & \ce{HCF}  & \ce{H2O} & \ce{BeH} & \ce{NH2} & \ce{HCO} \\
acetaldehyde & \ce{HCCl} & \ce{H2S} & \ce{BeF} & \ce{PH2} & \ce{HOC} \\
silylidene   & \ce{HPO}  & \ce{N2}  & \ce{BH2} & vinyl    & \ce{OH}  \\
nitroxyl     & \ce{CF2}  & \ce{NH3} & \ce{CN}  & allyl    & \ce{CO+} \\
ethylene     & \ce{BH}   & \ce{HCl} & \ce{CH3} &          & \\
methanimine  & \ce{BF}   &          &          &          & \\
\end{tabular}
\end{ruledtabular}
\end{table}

@@

We performed calculations following both the standard ground-state-based CI route and the state-specific CI route. \cite{Kossoski_2023}
For the latter, we employed the state-specific orbitals obtained in Ref.~\onlinecite{Kossoski_2023}.
Notice that, in contrast to eCI, the energies obtained with sCI and hCI models are not invariant under rotations within the occupied and virtual subspaces.
The restricted HF solution (restricted open-shell HF for the open-shell systems) was taken as the reference determinant for the hCI and eCI ground-state calculations
(left and middle panels of Fig.~\ref{fig:hCI}).
For the state-specific hCI and eCI calculations, we employed a minimal CSF reference: \cite{Kossoski_2023}
a single open-shell determinant for the doublet excited states and a single open-shell CSF for the excited states from closed-shell systems
(middle and right panels of Fig.~\ref{fig:hCI}, respectively).
The computed excitation energies were benchmarked against the reference theoretical values provided in the QUEST database. \cite{Veril_2021}
\fk{For each excited state considered here, the reference value and the method used to obtain it (of high-order CC or extrapolated FCI quality) can be found in the {\SupInf}.}

@@

%%% FIG 3 %%%
\begin{figure*}%[h!]
\includegraphics[width=1.0\linewidth]{plot_all}
\caption{
Non-parallelity error (NPE), distance error, equilibrium distance, and vibrational frequency (or force constant), for \ce{OH}, \ce{CN}, vinyl, and \ce{H7},
as functions of the number of determinants ($\Ndet$), according to the hCI (green), eCI (red), and sCI (blue) routes,
with (light tone crosses) and without (dark tone circles) the standard EN2 perturbative correction.
\fk{Each point denotes one CI model, according to the sequences: HF, hCI1, hCI1.5, hCI2, etc. (green); HF, CIS, CISD, etc. (red); and sCI1, sCI3, and sCI5 (blue).}
The dashed lines represent the FCI results.}
\label{fig:plot_all}
\end{figure*}
%%% %%% %%%

%%% FIG 4 %%%
\begin{figure*}%[h!]
\includegraphics[width=1.0\linewidth]{plot_all_closed}
\caption{
Non-parallelity error (NPE), distance error, equilibrium distance, and vibrational frequency (or force constant), for \ce{HF}, \ce{F2}, ethylene, \ce{N2}, \ce{H4}, and \ce{H8}, as functions of the number of determinants ($\Ndet$), according to the hCI (green), eCI (red) and sCI (blue) routes,
with (light tone crosses) and without (dark tone circles) the standard EN2 perturbative correction.
\fk{Each point denotes one CI model, according to the sequences: HF, hCI1, hCI1.5, hCI2, etc. (green); HF, CIS, CISD, etc. (red); and sCI0, sCI2, and sCI4 (blue).}
The dashed lines represent the FCI results.}
\label{fig:plot_all_closed}
\end{figure*}
%%% %%% %%%

@@

The distance errors obtained with hCI and eCI are overall comparable for open-shells, whereas the former presents slightly better convergence for the closed-shell systems.
In comparison to eCI, hCI leads to either comparable or somewhat faster (for \ce{CN}) convergence of the equilibrium geometries for the open-shell systems.
Finally, hCI systematically outperforms eCI in the calculation of vibrational frequencies, except for \ce{H7}, where no big difference is found.
\fk{For the \ce{CN} radical, CIS, sCI1, and sCI3 produce crossings between ground and excited states around equilibrium, hence non-smooth adiabatic PECs
and somewhat less reliable equilibrium properties in these specific cases.}
Given the overall slight superiority of hCI for the equilibrium properties, it also manages to account for more dynamic correlation than eCI,
for both open-shell and closed-shell systems.

@@

\fk{
We further explored hCI to describe the automerization barrier in the ground state of cyclobutadiene,
which connects the two equivalent rectangular $D_{2h}$ equilibrium geometries through the square $D_{4h}$ transition state geometry. \cite{Bally_1980,Tallarico_1996}
This is a well-known and challenging problem, requiring high-level calculations to achieve quantitative values for the height of the barrier (see Ref.~\onlinecite{Monino_2022} and references therein).
Here we employ the geometries presented in Ref.~\onlinecite{Monino_2022}, the 6-31+G(d) basis set, and the frozen-core approximation.
We optimize the orbitals for two closed-shell determinants in the $D_{2h}$ geometry, and for two spin-flipped open-shell determinants in the $D_{4h}$ geometry,
from which the excited determinants were generated in the subsequent CI calculations.
Even though the calculations target the ground state, they are labelled $\Delta$CI because different references are employed for each geometry.
For this basis set, the reference value of \SI{7.51}{kcal/mol} was obtained at CC with single, double, triple, and quadruple excitations. \cite{Monino_2022}
With respect to this reference value, the errors for the automerization barrier are
\SI{+3.55}{kcal/mol} with our two-determinant calculations, \SI{-2.28}{kcal/mol} with $\Delta$hCI2, \SI{-1.07}{kcal/mol} with $\Delta$hCI2+EN2, 
\SI{+0.61}{kcal/mol} with $\Delta$CISD, and \SI{-0.49}{kcal/mol} with $\Delta$CISD+EN2.
These values are comparable to those obtained with other methods, extensively discussed in Ref.~\onlinecite{Monino_2022}.
Here we just mention the errors of CC with singles and doubles, \SI{+0.80}{kcal/mol}, and of spin-flip equation-of-motion CC (EOM-CC) with singles and doubles, \SI{-1.65}{kcal/mol},
which share the same $\order*{N^6}$ computational scaling as $\Delta$CISD and $\Delta$hCI2.
This specific example serves to illustrate that the accuracy of hCI models can be similar to that of other methods.
As such, hCI can be taken into consideration in future systematic studies that tackle challenging chemical problems where strong correlation comes into play.
}

@@

For each CI model considered here, we evaluate the mean signed error (MSE), mean absolute error (MAE), root-mean-square error (RMSE), and standard deviation of the errors (SDE).
with respect to the reference theoretical values for the excitation energies.
\fk{For completeness, the defition of these statistical measures can be found in the {\SupInf}.}
For the lower-order models, the calculations were performed for 50 (closed-shell) and 19 (open-shell) excited states.
In turn, subsets of 16 (closed-shell) and 6 (open-shell) excited states were considered for the higher-order CI models, given the more intensive computational cost.
Even though these subsets are too small for meaningful absolute statistics, they should be enough to reveal main trends.
A detailed comparison of eCI based on ground-state and state-specific approaches can be found elsewhere. \cite{Kossoski_2023}
The focus of the present discussion lies on the comparison between hCI and eCI, and sCI,
and on a similar comparison between ground-state and state-specific approaches for hCI and sCI.
\fk{The bare (unperturbed) hCI and sCI models are discussed in this section and in Sec.~\ref{sec:res_D}, respectively,
whereas their EN2 perturbed analogues are left for Sec.~\ref{sec:res_E}.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\fk{hCI and sCI plus EN2 correction for excited states}}
\label{sec:res_E}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

% Abstract
\begin{abstract}
Hierarchy configuration interaction (hCI) has been recently introduced as an alternative configuration interaction (CI) route combining excitation degree and seniority number, which showed to efficiently recover both dynamic and static correlations for closed-shell molecular systems [\href{https://doi.org/10.1021/acs.jpclett.2c00730}{\textit{J.~Phys.~Chem.~Lett.}~\textbf{2022}, \textit{13}, 4342}].
Here, we generalize hCI for an arbitrary reference determinant, allowing calculations for radicals and for excited states in a state-specific way.
We gauge this route against excitation-based CI (eCI) and seniority-based CI (sCI) by evaluating how different ground-state properties of radicals converge to the full CI limit.
We find that hCI outperforms or matches eCI, whereas sCI is far less accurate, in line with previous observations for closed-shell molecules.
Employing the second-order Epstein-Nesbet (EN2) perturbation theory as a correction significantly accelerates the convergence of hCI and eCI.
We further explore various hCI and sCI models to calculate excitation energies of closed- and open-shell systems.
Our results underline that both the choice of the reference determinant and the set of orbitals drive the fine balance between correlation of ground and excited states.
State-specific hCI2 and higher order models perform similarly to their eCI counterparts, whereas lower orders of hCI deliver poor results,
unless supplemented by the EN2 correction, which substantially improves their accuracy.
In turn, sCI1 produces decent excitation energies for radicals, encouraging the development of related seniority-based coupled-cluster methods.
\bigskip
\begin{center}
        \boxed{\includegraphics[keepaspectratio,width=3.25in]{TOC}}
\end{center}
\bigskip
\end{abstract}

@@

Here, our first goal is to generalize hCI for an arbitrary type of reference, thus extending its applicability from ground-state closed-shells \cite{Kossoski_2022}
to radicals and state-specific excited-state calculations.
This is done in Sec.~\ref{sec:hCI}.
The computational details about the implementation and the specific calculations performed here are presented in Sec.~\ref{sec:compdet}.
In Sec.~\ref{sec:res_A}, we assess the performance of hCI, eCI, and sCI models by calculating various properties of four ground-state radicals, which comprises our second goal.
Our third goal is to evaluate how perturbation theory, more precisely the second-order Epstein-Nesbet (EN2) perturbative correction (both standard and renormalized), \cite{Garniron_2019}
impacts various properties of ground-state closed-shell systems and radicals, for hCI, eCI, and sCI models.
This part is presented in Sec.~\ref{sec:res_B}.
Inspired by the results of hCI for ground-state closed-shell systems, \cite{Kossoski_2022}
and by the promising set of excitation-based $\Delta$CI, \cite{Kossoski_2023} here we explore hCI models for excited states, following both the ground-state-based and state-specific approaches.
In this sense, our fourth goal is to assess the accuracy of hCI models for excited states, which is the subject of Sec.~\ref{sec:res_C}.
Furthermore, to the best of our knowledge, sCI models have not yet been directly used to target excited states,
despite the growing number of methods that exploit the concept of seniority,
for both ground
\cite{Limacher_2013,Limacher_2014,Tecmer_2014,Boguslawski_2014a,Boguslawski_2015,Boguslawski_2014b,Boguslawski_2014c,Johnson_2020,Henderson_2014,Stein_2014,Henderson_2015,Chen_2015,Bytautas_2018,Marie_2021,Boguslawski_2021,Tecmer_2022,Mamache_2023,Fecteau_2022}
and excited states.
\cite{Boguslawski_2016b,Boguslawski_2016c,Boguslawski_2019,Nowak_2019,Kossoski_2021,Marie_2021,Tecmer_2022,Rishi_2023,Nowak_2023,Fecteau_2022}
Our fifth goal, detailed in Sec.~\ref{sec:res_D}, is therefore to define and gauge ground-state-based and state-specific sCI models for excited states.
Finally, in Sec.~\ref{sec:res_E}, we assess how the excitation energies are impacted by the EN2 perturbative correction to hCI and sCI models, our sixth and last goal.
Section \ref{sec:conclusion} closes the present contribution with the main conclusions and perspectives.

@@

We introduced hCI \cite{Kossoski_2022} as a particular truncation of the Hilbert space,
viewed as a two-dimensional map of determinants built from their seniority $s$ and their excitation degree $e$ with respect to a reference determinant, as shown in Fig.~\ref{fig:hCI}.
By defining orbital subspaces according to the occupancy in the reference determinant (doubly-occupied, singly-occupied, or unoccupied),
we refer to the class of an excited determinant as the combination of the number of electrons and seniority in each orbital subspace.
The number of determinants $\Ndet$ within a given class scales polynomially with the number of basis functions $N$, with the exponent depending on the specific class.
For instance, for the class of doubly-excited determinants with no unpaired electrons ($e=2$ and $s=0$), $\Ndet = \order*{N^2}$.
hCI was defined such that, at a given hierarchy $h$, all classes of determinants presenting a scaling of $\Ndet = \order*{N^{2h}}$ or less are accounted for.
This means moving diagonally in the seniority-excitation map (denoted by the color tones in Fig.~\ref{fig:hCI}).
In comparison, eCI spans the map horizontally (top to bottom), whereas sCI does it vertically (left to right).

@@

The various CI models introduced here were further assessed based on calculated vertical excitation energies for 69 electronic states,
from 17 closed-shell systems and 13 radicals (shown in Table~\ref{tab:systems}), with geometries extracted from the QUEST database. \cite{Veril_2021}
Our set includes mostly small systems, ranging from \ce{BH} to glyoxal, displaying a mix of valence and Rydberg singly-excited states,
and 2 doubly-excited states (glyoxal and nitroxyl). It does not include large molecules nor charge transfer states.
The full set of excited states and calculated excitation energies, for the various CI models, are provided in the {\SupInf}.

\begin{table}[ht!]
\caption{Systems Considered in the Excited State Calculations.}
\label{tab:systems}
\begin{ruledtabular}
\begin{tabular}{llllll}
\mc{3}{c}{Closed-shells}           & \mc{3}{c}{Open-shells} \\
\cline{1-3} \cline{4-6}
glyoxal      & \ce{HCF}  & \ce{H2O} & \ce{BeH} & \ce{NH2} & \ce{HCO} \\
acetaldehyde & \ce{HCCl} & \ce{H2S} & \ce{BeF} & \ce{PH2} & \ce{HOC} \\
silylidene   & \ce{HPO}  & \ce{N2}  & \ce{BH2} & vinyl    & \ce{OH}  \\
nitroxyl     & \ce{CF2}  & \ce{NH3} & \ce{CN}  & allyl    & \ce{CO+} \\
ethylene     & \ce{BH}   & \ce{HCl} & \ce{CH3} &          & \\
methanimine  & \ce{BF}   &          &          &          & \\
\end{tabular}
\end{ruledtabular}
\end{table}

@@

We performed calculations following both the standard ground-state-based CI route and the state-specific CI route. \cite{Kossoski_2023}
For the latter, we employed the state-specific orbitals obtained in Ref.~\onlinecite{Kossoski_2023}.
Notice that, in contrast to eCI, the energies obtained with sCI and hCI models are not invariant under rotations within the occupied and virtual subspaces.
The restricted HF solution (restricted open-shell HF for the open-shell systems) was taken as the reference determinant for the hCI and eCI ground-state calculations
(left and middle panels of Fig.~\ref{fig:hCI}).
For the state-specific hCI and eCI calculations, we employed a minimal CSF reference: \cite{Kossoski_2023}
a single open-shell determinant for the doublet excited states and a single open-shell CSF for the excited states from closed-shell systems
(middle and right panels of Fig.~\ref{fig:hCI}, respectively).
The computed excitation energies were benchmarked against the reference theoretical values provided in the QUEST database. \cite{Veril_2021}
For each excited state considered here, the reference value and the method used to obtain it (of high-order CC or extrapolated FCI quality) can be found in the {\SupInf}.

@@

%%% FIG 3 %%%
\begin{figure*}%[h!]
\includegraphics[width=1.0\linewidth]{plot_all}
\caption{
Non-parallelity error (NPE), distance error, equilibrium distance, and vibrational frequency (or force constant), for \ce{OH}, \ce{CN}, vinyl, and \ce{H7},
as functions of the number of determinants ($\Ndet$), according to the hCI (green), eCI (red), and sCI (blue) routes,
with (light tone crosses) and without (dark tone circles) the standard EN2 perturbative correction.
Each point denotes one CI model, according to the sequences: HF, hCI1, hCI1.5, hCI2, etc. (green); HF, CIS, CISD, etc. (red); and sCI1, sCI3, and sCI5 (blue).
The dashed lines represent the FCI results.}
\label{fig:plot_all}
\end{figure*}
%%% %%% %%%

%%% FIG 4 %%%
\begin{figure*}%[h!]
\includegraphics[width=1.0\linewidth]{plot_all_closed}
\caption{
Non-parallelity error (NPE), distance error, equilibrium distance, and vibrational frequency (or force constant), for \ce{HF}, \ce{F2}, ethylene, \ce{N2}, \ce{H4}, and \ce{H8}, as functions of the number of determinants ($\Ndet$), according to the hCI (green), eCI (red) and sCI (blue) routes,
with (light tone crosses) and without (dark tone circles) the standard EN2 perturbative correction.
Each point denotes one CI model, according to the sequences: HF, hCI1, hCI1.5, hCI2, etc. (green); HF, CIS, CISD, etc. (red); and sCI0, sCI2, and sCI4 (blue).
The dashed lines represent the FCI results.}
\label{fig:plot_all_closed}
\end{figure*}
%%% %%% %%%

@@

The distance errors obtained with hCI and eCI are overall comparable for open-shells, whereas the former presents slightly better convergence for the closed-shell systems.
In comparison to eCI, hCI leads to either comparable or somewhat faster (for \ce{CN}) convergence of the equilibrium geometries for the open-shell systems.
Finally, hCI systematically outperforms eCI in the calculation of vibrational frequencies, except for \ce{H7}, where no big difference is found.
For the \ce{CN} radical, CIS, sCI1, and sCI3 produce crossings between ground and excited states around equilibrium, hence non-smooth adiabatic PECs
and somewhat less reliable equilibrium properties in these specific cases.
Given the overall slight superiority of hCI for the equilibrium properties, it also manages to account for more dynamic correlation than eCI,
for both open-shell and closed-shell systems.

@@

We further explored hCI to describe the automerization barrier in the ground state of cyclobutadiene,
which connects the two equivalent rectangular $D_{2h}$ equilibrium geometries through the square $D_{4h}$ transition state geometry. \cite{Bally_1980,Tallarico_1996}
This is a well-known and challenging problem, requiring high-level calculations to achieve quantitative values for the height of the barrier (see Ref.~\onlinecite{Monino_2022} and references therein).
Here we employ the geometries presented in Ref.~\onlinecite{Monino_2022}, the 6-31+G(d) basis set, and the frozen-core approximation.
We optimize the orbitals for two closed-shell determinants in the $D_{2h}$ geometry, and for two spin-flipped open-shell determinants in the $D_{4h}$ geometry,
from which the excited determinants were generated in the subsequent CI calculations.
Even though the calculations target the ground state, they are labelled $\Delta$CI because different references are employed for each geometry.
For this basis set, the reference value of \SI{7.51}{kcal/mol} was obtained at CC with single, double, triple, and quadruple excitations. \cite{Monino_2022}
With respect to this reference value, the errors for the automerization barrier are
\SI{+3.55}{kcal/mol} with our two-determinant calculations, \SI{-2.28}{kcal/mol} with $\Delta$hCI2, \SI{-1.07}{kcal/mol} with $\Delta$hCI2+EN2, 
\SI{+0.61}{kcal/mol} with $\Delta$CISD, and \SI{-0.49}{kcal/mol} with $\Delta$CISD+EN2.
These values are comparable to those obtained with other methods, extensively discussed in Ref.~\onlinecite{Monino_2022}.
Here we just mention the errors of CC with singles and doubles, \SI{+0.80}{kcal/mol}, and of spin-flip equation-of-motion CC (EOM-CC) with singles and doubles, \SI{-1.65}{kcal/mol},
which share the same $\order*{N^6}$ computational scaling as $\Delta$CISD and $\Delta$hCI2.
This specific example serves to illustrate that the accuracy of hCI models can be similar to that of other methods.
As such, hCI can be taken into consideration in future systematic studies that tackle challenging chemical problems where strong correlation comes into play.

@@

For each CI model considered here, we evaluate the mean signed error (MSE), mean absolute error (MAE), root-mean-square error (RMSE), and standard deviation of the errors (SDE).
with respect to the reference theoretical values for the excitation energies.
For completeness, the defition of these statistical measures can be found in the {\SupInf}.
For the lower-order models, the calculations were performed for 50 (closed-shell) and 19 (open-shell) excited states.
In turn, subsets of 16 (closed-shell) and 6 (open-shell) excited states were considered for the higher-order CI models, given the more intensive computational cost.
Even though these subsets are too small for meaningful absolute statistics, they should be enough to reveal main trends.
A detailed comparison of eCI based on ground-state and state-specific approaches can be found elsewhere. \cite{Kossoski_2023}
The focus of the present discussion lies on the comparison between hCI and eCI, and sCI,
and on a similar comparison between ground-state and state-specific approaches for hCI and sCI.
The bare (unperturbed) hCI and sCI models are discussed in this section and in Sec.~\ref{sec:res_D}, respectively,
whereas their EN2 perturbed analogues are left for Sec.~\ref{sec:res_E}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{hCI and sCI plus EN2 correction for excited states}
\label{sec:res_E}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The renormalized EN2 correction significantly reduces the errors across most CI models, for both closed- and open-shell systems.
The statistical measures for the CI models corrected with the EN2 correction can be seen in
Table~\ref{tab:stat1} (closed-shell systems) and Table~\ref{tab:stat2} (open-shell systems),
along with the results for the bare (unperturbed) CI models.
We have not considered the EN2 correction for the higher-order models.
